---
title: "Untitled"
output: html_document
date: "2025-01-13"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
options(dplyr.summarise.inform = FALSE)
options(scipen = 999)

library(tidyverse)
library(knitr)
library(scales)
library(kableExtra)
library(lubridate)
source("R/translate_table.R")
```

## Intro

This is for surveys delivered with a trip table (and not a place table).

## Translate Variables

```{r}
hh <- read_csv("data/_private/trm_survey/input/hh.csv")
per <- read_csv("data/_private/trm_survey/input/person.csv")
trip <- read_csv("data/_private/trm_survey/input/trip.csv")
loc <- read_csv("data/_private/trm_survey/input/location.csv")
```


```{r}
equiv_tbl <- read_csv("data/input/trm_survey/trm_equiv_table.csv") %>%
  mutate(VARIABLE = tolower(VARIABLE))
translate_hh <- translate_table(hh, equiv_tbl)
translate_per <- translate_table(per, equiv_tbl)
translate_trip <- translate_table(trip, equiv_tbl)
```

# Standardize fields

```{r}
hh_standard <- translate_hh %>%
  mutate(
    hh_id = hh_id,
    income_midpt = income_detailed,
    income_midpt = ifelse(is.na(income_midpt), income_broad, income_midpt),
    hhsize = num_people,
    hh_weight = hh_weight
  )

per_standard <- translate_per %>%
  mutate(
    hh_id = hh_id,
    person_id = person_id,
    employed = ifelse(employment == "unemployed", "No", "Yes"),
    age2 = ifelse(age_detailed == 999, 45, age_detailed),
    age = ifelse(is.na(age2), age, age2),
    license = can_drive
  ) %>%
  select(-age2)

trip_standard <- translate_trip %>%
  mutate(
    hh_id = hh_id,
    person_id = person_id,
    trip_num = trip_num,
    trip_id = trip_id,
    party = num_travelers,
    HH_Members = num_hh_travelers,
    Non_HH_Members = num_non_hh_travelers,
    arr_hr = arrive_hour,
    arr_min = arrive_minute,
    dep_hr = depart_hour,
    dep_min = depart_minute,
    activity_duration = duration_minutes,
    mode = mode_type,
    orig_purp = o_purpose_category,
    dest_purp = d_purpose_category,
    trip_weight = trip_weight
  )
```


# Geocoding

```{r, include=FALSE}
# Use this table in TransCAD to geocode points to TAZs
home_pts <- hh_standard %>%
  select(lon = home_lon, lat = home_lat)
work_pts <- per_standard %>%
  select(lon = work_lon, lat = work_lat)
school_pts <- per_standard %>%
  select(lon = school_lon, lat = school_lat)
o_pts <- trip_standard %>%
  select(lon = o_lon, lat = o_lat)
d_pts <- trip_standard %>%
  select(lon = d_lon, lat = d_lat)

to_geocode <- bind_rows(home_pts, work_pts, school_pts, o_pts, d_pts) %>%
  mutate(unique_id = paste0(lon, "_", lat)) %>%
  group_by(unique_id) %>%
  slice(1)
write_csv(to_geocode, "data/_private/trm_survey/output/to_geocode.csv")

# once geocoded externally, read it back in
geocoded_pts <- read_csv("data/_private/trm_survey/output/geocoded_points.csv") %>%
  select(unique_id, TAZ)

hh_with_tazs <- hh_standard %>%
  mutate(unique_id = paste0(home_lon, "_", home_lat)) %>%
  left_join(geocoded_pts, by = "unique_id") %>%
  rename(HomeTAZ = TAZ) %>%
  select(-unique_id)

per_with_taz <- per_standard %>%
  mutate(unique_id = paste0(work_lon, "_", work_lat)) %>%
  left_join(geocoded_pts, by = "unique_id") %>%
  rename(WorkTAZ = TAZ) %>%
  mutate(unique_id = paste0(school_lon, "_", school_lat)) %>%
  left_join(geocoded_pts, by = "unique_id") %>%
  rename(SchoolTAZ = TAZ) %>%
  select(-unique_id)


trip_with_taz <- trip_standard %>%
  select(-o_taz, -d_taz) %>%
  mutate(unique_id = paste0(o_lon, "_", o_lat)) %>%
  left_join(geocoded_pts, by = "unique_id") %>%
  rename(o_taz = TAZ) %>%
  mutate(unique_id = paste0(d_lon, "_", d_lat)) %>%
  left_join(geocoded_pts, by = "unique_id") %>%
  rename(d_taz = TAZ) %>%
  select(-unique_id)
```

## Create Trip File

Optional step if the survey came with a place file that needs to be converted
into a trip file.

```{r}
# trip_file <- remove_xfers %>%
#   group_by(hh_id, person_id) %>%
#   arrange(hh_id, person_id, trip_num) %>%
#   transmute(
#     trip_number = lead(trip_num),
#     dep_hr = dep_hr,
#     dep_min = dep_min,
#     arr_hr = lead(arr_hr),
#     arr_min = lead(arr_min),
#     activity_duration = lead(activity_duration),
#     mode = lead(mode_model),
#     HH_Members = lead(HH_Members),
#     Non_HH_Members = lead(Non_HH_Members),
#     orig_purp = dest_purp,
#     dest_purp = lead(dest_purp),
#     o_taz = TAZ,
#     d_taz = lead(TAZ)
#   ) %>%
#   filter(!is.na(dep_hr) & !is.na(arr_hr))

trip_file <- trip_with_taz
```

## Convert auto mode to sov/hov

```{r}
convert_modes <- trip_file %>%
  mutate(
    mode_model = case_when(
      mode == "auto" & party == 1 ~ "sov",
      mode == "auto" & party == 2 ~ "hov2",
      mode == "auto" & party >= 3 ~ "hov3",
      TRUE ~ mode
    )
  )
```

## Identify tours

```{r}
id_tours <- convert_modes %>%
  arrange(hh_id, person_id, trip_num) %>%
  group_by(hh_id, person_id) %>%
  mutate(
    tour_start = case_when(
      dest_purp == "H" & trip_num == max(trip_num) ~ 0,
      trip_num == min(trip_num) ~ 1,
      dest_purp == "H" ~ 1,
      TRUE ~ 0
    ),
    tour_num = cumsum((tour_start))
  ) %>%
  group_by(hh_id, person_id, tour_num) %>%
  mutate(
    tour_type = case_when(
      "W" %in% dest_purp ~ "work",
      TRUE ~ "other"
    ),
    tour_id = paste(hh_id, person_id, tour_num, sep = "_")
  )

# determine tour anchor place using the max duration, but break ties by
# purpose. For any remaining ties (same purpose and duration) just make the
# first one the anchor.
mark_anchor <- id_tours %>%
  group_by(hh_id, person_id) %>%
  mutate(
    activity_arr_mins = arr_hr * 60 + arr_min,
    activity_dep_mins = dep_hr * 60 + dep_min,
    # For people working 3rd shift, their last location is not home and does
    # not have a departure time. Give that a time of midnight for anchor calcs.
    activity_dep_mins = ifelse(
      dest_purp != "H" & trip_num == max(trip_num),
      24 * 60, activity_dep_mins
    ),
    activity_duration = ifelse(
      activity_arr_mins < activity_dep_mins,
      activity_dep_mins - activity_arr_mins,
      activity_dep_mins + 24 * 60 - activity_arr_mins
    ),
    activity_duration = ifelse(is.na(activity_duration), 0, activity_duration),
    activity_duration = as.integer(activity_duration),
    temp_dur = ifelse(dest_purp == "H", 0, activity_duration)
  ) %>%
  group_by(hh_id, person_id, tour_num) %>%
  mutate(
    anchor = ifelse(temp_dur == max(temp_dur), 100, 0),
    anchor = case_when(
      dest_purp == "W" ~ anchor + 3,
      dest_purp == "U" ~ anchor + 2,
      dest_purp == "Sch" ~ anchor + 1,
      TRUE ~ anchor,
    ),
    anchor = ifelse(anchor == max(anchor), 1, 0),
    check = cumsum(anchor),
    anchor = ifelse(check > 1, 0, anchor),
    anchor_taz = ifelse(anchor == 1, d_taz, NA),
    anchor_duration = ifelse(anchor == 1, temp_dur, NA),
    # Calculate intermediate stops
    int_stop = ifelse(
      trip_num == min(trip_num) | 
        trip_num == max(trip_num) | anchor == 1,
      0, 1
    ),
    int_stop_fwd = ifelse(check == 0, int_stop, 0), # before anchor
    int_stop_ret = ifelse(check >= 1, int_stop, 0)  # after anchor
  ) %>%
  select(-temp_dur, -check)

# Identify primary and secondary modes. Primary is the highest-order mode on
# trip. The secondary mode is the second highest order.
id_tour_mode <- mark_anchor %>%
  mutate(
    primary_mode = case_when(
      "airplane" %in% mode_model ~ "airplane",
      "rail" %in% mode_model ~ "rail",
      "school_bus" %in% mode_model ~ "school_bus",
      "bus" %in% mode_model ~ "bus",
      "tnc" %in% mode_model ~ "tnc",
      "hov3" %in% mode_model ~ "hov3",
      "hov2" %in% mode_model ~ "hov2",
      "sov" %in% mode_model ~ "sov",
      "bike" %in% mode_model ~ "bike",
      "walk" %in% mode_model ~ "walk",
      "other" %in% mode_model ~ "other",
      TRUE ~ "other"
    ),
    temp = ifelse(mode_model == primary_mode, NA, mode_model),
    secondary_mode = case_when(
      "airplane" %in% mode_model ~ "airplane",
      "rail" %in% temp ~ "rail",
      "school_bus" %in% temp ~ "school_bus",
      "bus" %in% temp ~ "bus",
      "tnc" %in% temp ~ "tnc",
      "hov3" %in% temp ~ "hov3",
      "hov2" %in% temp ~ "hov2",
      "sov" %in% temp ~ "sov",
      "bike" %in% temp ~ "bike",
      "walk" %in% temp ~ "walk",
      "other" %in% temp ~ "other",
      TRUE ~ primary_mode
    )
  ) %>%
  select(-temp)
```

## Add fields to trip file

```{r}
trip_add_fields <- id_tour_mode %>%
  mutate(
    pa_format = ifelse(dest_purp == "H", 0, 1),
    p_taz = ifelse(pa_format == 1, o_taz, d_taz),
    a_taz = ifelse(pa_format == 1, d_taz, o_taz),
    p_purp = ifelse(pa_format == 1, orig_purp, dest_purp),
    a_purp = ifelse(pa_format == 1, dest_purp, orig_purp),
    trip_purpose = case_when(
      tour_type == "work" & p_purp == "H" & a_purp == "W" ~ "W_HBW",
      tour_type == "work" & p_purp == "H" ~ "W_HBO",
      tour_type == "work" ~ "W_NHB",
      tour_type != "work" & p_purp == "H" & a_purp == "Sch" ~ "N_HBSch",
      tour_type != "work" & p_purp == "H" & activity_duration < 30 ~ "N_HBOS",
      tour_type != "work" & p_purp == "H" ~ "N_HBOL",
      TRUE ~ "N_NHB"
    ),
    ix_type = case_when(
      is.na(o_taz) & is.na(d_taz) ~ "EE",
      !is.na(o_taz) & !is.na(d_taz) ~ "II",
      TRUE ~ "IX"
    )
  )

# Add TOD data
trip_add_tod <- trip_add_fields %>%
  mutate(
    trip_dep_mins = dep_hr * 60 + dep_min,
    trip_arr_mins = arr_hr * 60 + arr_min,
    midpoint = (trip_arr_mins - trip_dep_mins) / 2 + trip_dep_mins,
    tod = case_when(
      trip_dep_mins < 420 ~ "NT",
      trip_dep_mins < 540 ~ "AM",
      trip_dep_mins < 900 ~ "MD",
      trip_dep_mins < 1080 ~ "PM",
      TRUE ~ "NT",
    )) %>%
  select(-trip_dep_mins, trip_arr_mins, midpoint)

# Add skim data
skim <- read_csv("data/input/skims/skim.csv") %>%
  rename(skim_length = Length, skim_time = Time)
trip_add_skim <- trip_add_tod %>%
  left_join(skim, by = c("p_taz" = "Origin", "a_taz" = "Destination"))
```

# Create Tours/Stops

```{r}
tour_file <- trip_add_skim %>%
  group_by(tour_id) %>%
  filter(!all(is.na(anchor_taz))) %>%
  summarize(
    hh_id = first(hh_id),
    person_id = first(person_id),
    tour_num = first(tour_num),
    tour_type = first(tour_type),
    num_trips = n(),
    anchor_taz = max(anchor_taz, na.rm = TRUE),
    anchor_start_hr = sum(arr_hr[anchor == 1]),
    anchor_start_min = sum(arr_min[anchor == 1]),
    anchor_end_hr = sum(dep_hr[anchor == 1]),
    anchor_end_min = sum(dep_min[anchor == 1]),
    anchor_duration = max(anchor_duration, na.rm = TRUE),
    int_stop_fwd = sum(int_stop_fwd),
    int_stop_ret = sum(int_stop_ret),
    primary_mode = first(primary_mode),
    secondary_mode = first(secondary_mode),
    HH_Members_Max = max(HH_Members, na.rm = TRUE),
    HH_Members_Min = min(HH_Members, na.rm = TRUE),
    Non_HH_Members_Max = max(Non_HH_Members, na.rm = TRUE),
    Non_HH_Members_Min = min(Non_HH_Members, na.rm = TRUE)
  ) %>%
  mutate(
    anchor_taz = ifelse(anchor_taz == -Inf, NA, anchor_taz),
    # members max/min can be inf/-inf. This is mostly for home tours
    # (people who make no tours), but it also affects open tours at the start
    # of the day. Set them all to 1 person.
    across(HH_Members_Max:HH_Members_Min, ~ ifelse(is.finite(.x), .x, 1)),
    across(Non_HH_Members_Max:Non_HH_Members_Min, ~ ifelse(is.finite(.x), .x, 0))
  ) %>%
  ungroup()

stops_file <- trip_add_skim %>%
  ungroup() %>%
  filter(int_stop == 1) %>%
  mutate(stop_taz = d_taz) %>%
  select(
    tour_id, int_stop_fwd, int_stop_ret, stop_taz,
    dest_purp, duration = activity_duration
  ) %>%
  # determine primary forward and return stop based on duration
  group_by(tour_id) %>%
  mutate(stop_id = row_number()) %>%
  ungroup() %>%
  arrange(tour_id, int_stop_fwd, desc(duration)) %>%
  group_by(tour_id, int_stop_fwd) %>%
  mutate(primary_stop = ifelse(row_number() == 1, 1, 0)) %>%
  ungroup() %>%
  arrange(tour_id, stop_id) %>%
  relocate(stop_id, .after = tour_id)
```


# Household Processing

## Calculate HHStructure

```{r}
per_calc_ptypes <- per_with_taz %>%
  mutate(
    WAdult = ifelse(employed == "Yes" & age >= 18, 1, 0),
    NWAdult = ifelse(employed == "No" & age >= 18, 1, 0),
    Kid = ifelse(age < 18, 1, 0),
    PreSch = ifelse(age < 5, 1, 0),
    Senior = ifelse(age >= 65, 1, 0),
    Licensed_Driver = ifelse(license == "Yes", 1, 0)
  )

hhstructure <- per_calc_ptypes %>%
  group_by(hh_id) %>%
  summarise(
    WAdults = sum(WAdult), NWAdults = sum(NWAdult), HHKids = sum(Kid), 
    HHPreSchs = sum(PreSch), HHSeniors = sum(Senior),
    HHDrvLic = sum(Licensed_Driver)
  ) %>%
  mutate(HHStructure = paste0("HH_", WAdults, NWAdults, HHKids))

hh_add_strct <- hh_with_tazs %>%
  left_join(hhstructure, by = "hh_id")
  
```

```{r}
# Update household file with numbers of workers who took a work tour that day
tours_by_hh <- tour_file %>%
  group_by(hh_id, person_id, tour_type) %>%
  summarize(count = n()) %>%
  pivot_wider(
    id_cols = c(hh_id, person_id), names_from = tour_type, 
    values_from = count, values_fill = 0
  ) %>%
  mutate(WorkersThatDay = ifelse(work > 0, 1, 0)) %>%
  group_by(hh_id) %>%
  summarize(across(-person_id, ~ sum(.x))) %>%
  rename_with(~ paste0(.x, "_tours"), -c("hh_id", "WorkersThatDay"))
  

hh_add_tours <- hh_add_strct %>%
  left_join(tours_by_hh, by = "hh_id") %>%
  mutate(across(ends_with("_tours"), ~ifelse(is.na(.), 0, .))) %>%
  mutate(WorkersThatDay = ifelse(is.na(WorkersThatDay), 0, WorkersThatDay)) %>%
  relocate(WorkersThatDay, .after = WAdults) %>%
  mutate(
    BasicPattern = paste0(
      "P_",
      WorkersThatDay,
      pmax(hhsize - HHKids - WorkersThatDay, 0), # adults not working that day,
      HHKids
    )
  )
```

# Person Processing

## Add tour summary to person file

```{r}
tours_by_per <- tour_file %>%
  group_by(hh_id, person_id, tour_type) %>%
  summarize(count = n()) %>%
  pivot_wider(
    id_cols = hh_id:person_id, names_from = tour_type, 
    values_from = count, values_fill = 0
  ) %>%
  rename_with(~ paste0(.x, "_tours"), -c(hh_id, person_id))

per_add_tour_counts <- per_calc_ptypes %>%
  left_join(
    hh_add_strct %>% select(hh_id, year, HomeTAZ), by = c("hh_id", "year")
  ) %>%
  left_join(tours_by_per, by = c("hh_id", "person_id")) %>%
  mutate(across(ends_with("_tours"), ~ifelse(is.na(.), 0, .)))
```

```{r}
per_calc_iz <- per_add_tour_counts %>%
  mutate(
    IZ_Work = ifelse(HomeTAZ == WorkTAZ, 1, 0),
    IZ_School = ifelse(HomeTAZ == SchoolTAZ, 1, 0)
  )
```

# Handle weights/filtering

```{r}
# only keep households from 2022
hh_weights <- hh_add_tours %>% 
  mutate(hh_weight = ifelse(year == 2022, hh_weight, NA)) %>%
  filter(!is.na(hh_weight))

per_weights <- per_calc_iz %>% 
  left_join(hh_weights %>% select(hh_id, hh_weight), by = "hh_id") %>%
  filter(!is.na(hh_weight))

tour_weights <- tour_file %>% 
  left_join(hh_weights %>% select(hh_id, hh_weight), by = "hh_id") %>%
  filter(!is.na(hh_weight))

stop_weights <- stops_file %>%
  left_join(tour_weights %>% select(tour_id, hh_weight), by = "tour_id") %>%
  filter(!is.na(hh_weight))

trip_weights <- trip_add_skim %>% 
  left_join(hh_weights %>% select(hh_id, hh_weight), by = "hh_id") %>%
  filter(!is.na(hh_weight))
```

# Final tables

```{r}
# finalize columns
hh_final <- hh_weights %>%
  select(
    hh_id = hh_id,
    hh_weight,
    HomeTAZ,
    WAdults,
    WorkersThatDay,
    NWAdults,
    HHKids,
    HHPreSchs,
    HHSeniors,
    HHDrvLic,
    HHStructure,
    income_midpt,
    other_tours,
    work_tours,
    BasicPattern
  )

tour_final <- tour_weights %>%
  transmute(
    tour_id,
    hh_id = hh_id,
    person_id = person_id,
    tour_id,
    tour_type,
    num_trips,
    anchor_taz,
    anchor_start_hr,
    anchor_start_min,
    anchor_duration,
    int_stop_fwd,
    int_stop_ret,
    primary_mode,
    secondary_mode,
    HH_Members_Max,
    HH_Members_Min,
    Non_HH_Members_Max = NA,
    Non_HH_Members_Min = NA,
    hh_weight
  )

stop_final <- stop_weights %>%
  transmute(
    tour_id,
    stop_id,
    int_stop_fwd,
    int_stop_ret,
    stop_taz,
    a_purp = dest_purp,
    duration,
    primary_stop,
    hh_weight
  )

trip_final <- trip_weights %>%
  transmute(
    hh_id = hh_id,
    person_id = person_id,
    trip_id = trip_id,
    trip_weight = trip_weight,
    hh_weight,
    p_purp = p_purp,
    a_purp = a_purp,
    p_taz,
    a_taz,
    tour_id = tour_num,
    tour_type,
    trip_purpose,
    arrive_hour = arr_hr,
    arrive_minute = arr_min,
    duration_minutes = activity_duration,
    mode = mode,
    ix_type,
    skim_length,
    skim_time
  )
```


# Write out

```{r, eval=FALSE}
write_csv(hh_final, "data/_private/trm_survey/output/output_households.csv", na = "")
write_csv(tour_final, "data/_private/trm_survey/output/output_tours.csv", na = "")
write_csv(stop_final, "data/_private/trm_survey/output/output_stops.csv", na = "")
write_csv(trip_final, "data/_private/trm_survey/output/output_trips.csv", na = "")
```

