---
title: "Untitled"
output: html_document
date: "2025-01-13"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
options(dplyr.summarise.inform = FALSE)
options(scipen = 999)

library(tidyverse)
library(knitr)
library(scales)
library(kableExtra)
library(lubridate)
library(hms)
library(readxl)
source("R/translate_table.R")
```

## Intro

This is for surveys delivered with a trip table (and not a place table).

## Translate Variables

```{r}
per <- read_excel("data/_private/vis_survey_2024/Oahu 2024 Visitor Survey Database Final Version Submitted with data dictionary_Revised Feb 26 2025.xlsx", sheet = 2)
loc <- read_excel("data/_private/vis_survey_2024/Oahu 2024 Visitor Survey Database Final Version Submitted with data dictionary_Revised Feb 26 2025.xlsx", sheet = 4)
```


```{r}
equiv_tbl <- read_csv("data/input/vis_survey_2024/vis2024_equiv_table.csv")
# translate_hh <- translate_table(hh, equiv_tbl)
translate_per <- translate_table(per, equiv_tbl)
# translate_trip <- translate_table(trip, equiv_tbl)
translate_loc <- translate_table(loc, equiv_tbl)
```

# Standardize fields

```{r}
# simplify field names
clean_column_names <- function(df) {
  colnames(df) <- colnames(df) %>%
    tolower() %>%                     # Convert to lowercase
    gsub("[^a-z0-9 ]", "", .) %>%      # Remove special characters except spaces
    gsub("\\s+", "_", .)               # Replace spaces with underscores
  return(df)
}

per_simple <- clean_column_names(translate_per)
loc_simple <- clean_column_names(translate_loc)
```


```{r}
per_standard <- per_simple %>%
  mutate(
    # hh_id = hh_id,
    person_id = personid2,
    employed = ifelse(employment_status == "No", "No", "Yes"),
    age = what_is_your_age
  ) %>%
  rename(
    hh_income = household_income_code,
    prepay = did_you_prepay_for_any_transportation_services_such_as_trolley_tickets_tour_groups_etc_before_you_arrived_in_oahu,
    days_outside_oahu = how_many_days_did_you_stay_on_islands_outside_oahu_during_this_trip_to_hawaii,
    overnight_outside_oahu = did_you_stay_overnight_on_any_islands_outside_oahu_during_this_trip_to_hawaii,
    travel_other_islands = did_you_travel_to_any_islands_other_than_oahu_during_your_visit_to_hawaii,
    freq_pt_athome= how_often_do_you_use_public_transportation_bus_tram_metrosubway_train_etc_in_the_area_where_you_live,
    help_env = use_if_it_helped_with_environmental_and_energy_issues,
    dont_like_strangers = why_not_i_prefer_not_to_be_in_close_proximity_to_strangers,
    trouble_holocard = why_not_had_trouble_getting_a_holo_card_bus_card,
    too_complicated = why_not_understanding_how_to_use_the_bus_seemed_too_complicated,
    unfamiliar = why_not_lack_of_familiarity_with_public_transit_system,
    not_convenient = why_not_convenience_of_other_transportation_options,
    use_pt = during_your_stay_on_oahu_did_you_use_public_transit_services_bus_or_rail,
    how_from_airport = how_did_you_travel_from_the_airport_to_the_place_you_stayed_on_oahu,
    number_hh_members = how_many_of_these_people_are_members_of_your_household,
    party_size = including_you_how_many_people_traveled_with_you_on_this_trip_to_oahu,
    time_arrive = what_time_did_you_arrive_on_oahu,
    nights = how_many_nights_did_you_will_you_spend_on_oahu_during_this_visit,
    main_purpose = what_is_the_main_purpose_of_your_trip_to_oahu,
    how_many_visits = including_this_trip_how_many_times_have_you_visited_oahu
  ) %>%
  select(
    -did_you_prepay_for_any_transportation_services_such_as_trolley_tickets_tour_groups_etc_before_you_arrived_in_oahu_code,
    -did_you_stay_overnight_on_any_islands_outside_oahu_during_this_trip_to_hawaii_code,
    -did_you_travel_to_any_islands_other_than_oahu_during_your_visit_to_hawaii_code,
    -how_often_do_you_use_public_transportation_bus_tram_metrosubway_train_etc_in_the_area_where_you_livecode,
    -during_your_stay_on_oahu_did_you_use_public_transit_services_bus_or_railcode,
    -how_did_you_trvel_from_the_airport_to_the_plce_you_styed_on_oahu_code,
    -what_time_did_you_arrive_on_oahu_code,
    -what_is_the_main_purpose_of_your_trip_to_oahu_code
  )

loc_standard <- loc_simple %>%
  select(-mode, -purpose) %>%
  mutate(
    person_id = personid,
    trip_num = tripnumber,
    arr_time = parse_date_time(arrive_time, orders = "h:M p"),
    arr_hr = hour(arr_time),
    arr_min = minute(arr_time),
    dep_time = parse_date_time(depart_time, orders = "h:M p"),
    dep_hr = hour(dep_time),
    dep_min = minute(dep_time),
    mode = mode_code,
    party_size = how_many_others_traveled_with_you,
    purpose = purpose_code
  ) %>%
  rename(
    paid_fees = did_you_pay_any_fees_for_the_service_taxi_uber_etc_you_used
  ) %>%
  select(
    -did_you_pay_any_fees_for_the_service_taxi_uber_etc_you_used_code
  )
```


# Geocoding

```{r, include=FALSE}
# Use this table in TransCAD to geocode points to TAZs
per_pts <- per_standard %>%
  select(lon = place_stayed_lon, lat = place_stayed_lat)
loc_pts <- loc_standard %>%
  select(lon = destination_lon, lat = destination_lat)

to_geocode <- bind_rows(per_pts, loc_pts) %>%
  mutate(unique_id = paste0(lon, "_", lat)) %>%
  group_by(unique_id) %>%
  slice(1)
write_csv(to_geocode, "data/_private/vis_survey_2024/output/to_geocode.csv")

# once geocoded externally, read it back in
geocoded_pts <- read_csv(
  "data/_private/vis_survey_2024/output/geocoded_points.csv") %>%
  select(unique_id, TAZ)

per_with_taz <- per_standard %>%
  mutate(unique_id = paste0(place_stayed_lon, "_", place_stayed_lat)) %>%
  select(-taz) %>%
  left_join(geocoded_pts, by = "unique_id") %>%
  rename(HomeTAZ = TAZ) %>%
  select(-unique_id)

loc_with_taz <- loc_standard %>%
  select(-taz) %>%
  mutate(unique_id = paste0(destination_lon, "_", destination_lat)) %>%
  left_join(geocoded_pts, by = "unique_id") %>%
  # rename(d_taz = TAZ) %>%
  select(-unique_id)
```

## Create Trip File

Optional step if the survey came with a place/location file that needs to be 
converted into a trip file.

```{r}
trip_file <- loc_with_taz %>%
  group_by(person_id) %>%
  arrange(person_id, trip_num) %>%
  transmute(
    trip_num = lead(trip_num),
    dep_hr = dep_hr,
    dep_min = dep_min,
    arr_hr = lead(arr_hr),
    arr_min = lead(arr_min),
    dep_mins = dep_hr * 60 + dep_min,
    arr_mins = arr_hr * 60 + arr_min,
    activity_duration = lead(dep_mins) - arr_mins,
    mode = lead(mode),
    party_size = lead(party_size),
    orig_purp = purpose,
    dest_purp = lead(purpose),
    o_taz = TAZ,
    d_taz = lead(TAZ)
  ) %>%
  select(-dep_mins, -arr_mins) %>%
  filter(!is.na(dep_hr) & !is.na(arr_hr))
```

## Convert auto mode to sov/hov

```{r}
convert_modes <- trip_file %>%
  mutate(
    mode = case_when(
      mode == "auto" & party_size == 1 ~ "sov",
      mode == "auto" & party_size == 2 ~ "hov2",
      mode == "auto" & party_size >= 3 ~ "hov3",
      TRUE ~ mode
    )
  )
```

## Identify tours

```{r}
id_tours <- convert_modes %>%
  arrange(person_id, trip_num) %>%
  group_by(person_id) %>%
  mutate(
    tour_start = case_when(
      dest_purp == "H" & trip_num == max(trip_num) ~ 0,
      trip_num == min(trip_num) ~ 1,
      orig_purp == "H" ~ 1,
      TRUE ~ 0
    ),
    tour_num = cumsum((tour_start))
  ) %>%
  group_by(person_id, tour_num) %>%
  mutate(
    tour_type = case_when(
      "W" %in% dest_purp ~ "work",
      "Shop" %in% dest_purp ~ "shop",
      "Rec" %in% dest_purp ~ "rec",
      TRUE ~ "other"
    ),
    tour_id = paste(person_id, tour_num, sep = "_")
  )

# determine tour anchor place using the max duration, but break ties by
# purpose. For any remaining ties (same purpose and duration) just make the
# first one the anchor.
mark_anchor <- id_tours %>%
  group_by(person_id) %>%
  mutate(
    arr_mins = arr_hr * 60 + arr_min,
    dep_mins = dep_hr * 60 + dep_min,
    temp_dur = ifelse(dest_purp == "H", 0, activity_duration)
  ) %>%
  group_by(person_id, tour_num) %>%
  mutate(
    anchor = ifelse(temp_dur == max(temp_dur), 100, 0),
    anchor = case_when(
      dest_purp == "W" ~ anchor + 3,
      dest_purp == "U" ~ anchor + 2,
      dest_purp == "Sch" ~ anchor + 1,
      TRUE ~ anchor,
    ),
    anchor = ifelse(anchor == max(anchor), 1, 0),
    check = cumsum(anchor),
    anchor = ifelse(check > 1, 0, anchor),
    anchor_taz = ifelse(anchor == 1, d_taz, NA),
    anchor_duration = ifelse(anchor == 1, temp_dur, NA),
    # Calculate intermediate stops
    int_stop = ifelse(
      trip_num == min(trip_num) | 
        trip_num == max(trip_num) | anchor == 1,
      0, 1
    ),
    int_stop_fwd = ifelse(check == 0, int_stop, 0), # before anchor
    int_stop_ret = ifelse(check >= 1, int_stop, 0)  # after anchor
  ) %>%
  select(-temp_dur, -check)

# Identify primary and secondary modes. Primary is the highest-order mode on
# trip. The secondary mode is the second highest order.
id_tour_mode <- mark_anchor %>%
  mutate(
    primary_mode = case_when(
      "transit" %in% mode ~ "transit",
      "tnc" %in% mode ~ "tnc",
      "hov3" %in% mode ~ "hov3",
      "hov2" %in% mode ~ "hov2",
      "sov" %in% mode ~ "sov",
      "bike" %in% mode ~ "bike",
      "walk" %in% mode ~ "walk",
      "other" %in% mode ~ "other",
      TRUE ~ "other"
    ),
    temp = ifelse(mode == primary_mode, NA, mode),
    secondary_mode = case_when(
      "transit" %in% temp ~ "transit",
      "tnc" %in% temp ~ "tnc",
      "hov3" %in% temp ~ "hov3",
      "hov2" %in% temp ~ "hov2",
      "sov" %in% temp ~ "sov",
      "bike" %in% temp ~ "bike",
      "walk" %in% temp ~ "walk",
      "other" %in% temp ~ "other",
      TRUE ~ primary_mode
    )
  ) %>%
  select(-temp)
```

## Add fields to trip file

```{r}
trip_add_fields <- id_tour_mode %>%
  mutate(
    pa_format = ifelse(dest_purp == "H", 0, 1),
    p_taz = ifelse(pa_format == 1, o_taz, d_taz),
    a_taz = ifelse(pa_format == 1, d_taz, o_taz),
    p_purp = ifelse(pa_format == 1, orig_purp, dest_purp),
    a_purp = ifelse(pa_format == 1, dest_purp, orig_purp),
    trip_purpose = case_when(
      tour_type == "work" & p_purp == "H" & a_purp == "W" ~ "W_HBW",
      tour_type == "work" & p_purp == "H" ~ "W_HBO",
      tour_type == "work" ~ "W_NHB",
      tour_type != "work" & p_purp == "H" & a_purp == "Rec" ~ "N_HBRec",
      tour_type != "work" & p_purp == "H" & a_purp == "Shop" ~ "N_HBShop",
      tour_type != "work" & p_purp == "H" & a_purp == "Sch" ~ "N_HBSch",
      tour_type != "work" & p_purp == "H" ~ "N_HBO",
      TRUE ~ "N_NHB"
    ),
    ix_type = case_when(
      is.na(o_taz) & is.na(d_taz) ~ "EE",
      !is.na(o_taz) & !is.na(d_taz) ~ "II",
      TRUE ~ "IX"
    )
  )

# Add TOD data
trip_add_tod <- trip_add_fields %>%
  mutate(
    trip_dep_mins = dep_hr * 60 + dep_min,
    trip_arr_mins = arr_hr * 60 + arr_min,
    midpoint = (trip_arr_mins - trip_dep_mins) / 2 + trip_dep_mins,
    tod = case_when(
      trip_dep_mins < 360 ~ "NT",
      trip_dep_mins < 540 ~ "AM",
      trip_dep_mins < 900 ~ "MD",
      trip_dep_mins < 1140 ~ "PM",
      TRUE ~ "NT",
    )) %>%
  select(-trip_dep_mins, trip_arr_mins, midpoint)

# Add skim data
skim <- read_csv("data/input/skims/auto_skim_am.csv") %>%
  rename(skim_length = auto_distance, skim_time = auto_time)
trip_add_skim <- trip_add_tod %>%
  left_join(skim, by = c("p_taz", "a_taz"))
```

# Create Tours/Stops

```{r}
tour_file <- trip_add_skim %>%
  group_by(tour_id) %>%
  filter(!all(is.na(anchor_taz))) %>%
  summarize(
    person_id = first(person_id),
    tour_num = first(tour_num),
    tour_type = first(tour_type),
    num_trips = n(),
    anchor_taz = max(anchor_taz, na.rm = TRUE),
    anchor_start_hr = sum(arr_hr[anchor == 1]),
    anchor_start_min = sum(arr_min[anchor == 1]),
    anchor_end_hr = sum(dep_hr[anchor == 1]),
    anchor_end_min = sum(dep_min[anchor == 1]),
    anchor_duration = max(anchor_duration, na.rm = TRUE),
    int_stop_fwd = sum(int_stop_fwd),
    int_stop_ret = sum(int_stop_ret),
    primary_mode = first(primary_mode),
    secondary_mode = first(secondary_mode),
    party_size_max = max(party_size, na.rm = TRUE),
    party_size_min = min(party_size, na.rm = TRUE)
  ) %>%
  mutate(
    anchor_taz = ifelse(anchor_taz == -Inf, NA, anchor_taz),
    # members max/min can be inf/-inf. This is mostly for home tours
    # (people who make no tours), but it also affects open tours at the start
    # of the day. Set them all to 1 person.
    across(party_size_max:party_size_min, ~ ifelse(is.finite(.x), .x, 1)),
    across(party_size_max:party_size_min, ~ ifelse(is.finite(.x), .x, 0))
  ) %>%
  ungroup()

stops_file <- trip_add_skim %>%
  ungroup() %>%
  filter(int_stop == 1) %>%
  mutate(stop_taz = d_taz) %>%
  select(
    tour_id, int_stop_fwd, int_stop_ret, stop_taz,
    dest_purp, duration = activity_duration
  ) %>%
  # determine primary forward and return stop based on duration
  group_by(tour_id) %>%
  mutate(stop_id = row_number()) %>%
  ungroup() %>%
  arrange(tour_id, int_stop_fwd, desc(duration)) %>%
  group_by(tour_id, int_stop_fwd) %>%
  mutate(primary_stop = ifelse(row_number() == 1, 1, 0)) %>%
  ungroup() %>%
  arrange(tour_id, stop_id) %>%
  relocate(stop_id, .after = tour_id)
```

# Person Processing

## Add tour summary to person file

```{r}
tours_by_per <- tour_file %>%
  group_by(person_id, tour_type) %>%
  summarize(count = n()) %>%
  pivot_wider(
    id_cols = person_id, names_from = tour_type, 
    values_from = count, values_fill = 0
  ) %>%
  rename_with(~ paste0(.x, "_tours"), -c(person_id))

per_add_tour_counts <- per_with_taz %>%
  left_join(tours_by_per, by = "person_id") %>%
  mutate(across(ends_with("_tours"), ~ifelse(is.na(.), 0, .)))
```

# Handle weights/filtering

```{r}
per_weights <- per_add_tour_counts %>% 
  mutate(weight = weight_factor_to_avg_daily_visitors_per_weekday_on_oahu_in_july_2024) %>%
  select(-weight_factor_to_avg_daily_visitors_per_weekday_on_oahu_in_july_2024) %>%
  filter(weight != 0)

tour_weights <- tour_file %>% 
  left_join(per_weights %>% select(person_id, weight), by = "person_id") %>%
  filter(weight != 0)

stop_weights <- stops_file %>%
  left_join(tour_weights %>% select(tour_id, weight), by = "tour_id") %>%
  filter(!is.na(weight))

trip_weights <- trip_add_skim %>% 
  left_join(per_weights %>% select(person_id, weight), by = "person_id") %>%
  filter(weight != 0)
```

```{r}
per_weights %>%
  group_by(county_origin_group) %>%
  summarize(across(ends_with("_tours"), ~weighted.mean(., w = weight)))

per_weights %>%
  group_by(main_purpose_grouped) %>%
  summarize(across(ends_with("_tours"), ~weighted.mean(., w = weight)))
```

# Final tables

```{r}
# same as weight tables in this case
```


# Write out

```{r, eval=FALSE}
write_csv(per_weights, "data/_private/vis_survey_2024/output/output_persons.csv", na = "")
write_csv(tour_weights, "data/_private/vis_survey_2024/output/output_tours.csv", na = "")
write_csv(stop_weights, "data/_private/vis_survey_2024/output/output_stops.csv", na = "")
write_csv(trip_weights, "data/_private/vis_survey_2024/output/output_trips.csv", na = "")
```

